---
title: "Seeding As a Prediction of Final Four Appearance"
format: html
editor: visual
---

## Background

The men's NCAA Final Four in 2023 seems to have a preponderance of lower-than-expected seeds. Over time, have the seeds gotten better or worse at predicting who will land in the Final Four?

If worse, it seems like the transfer portal and NIL have had the side effect of driving more parity in D-1 basketball, which makes it harder for the seeds to reflect truly cut-above-the-rest top seeds.

## The Analysis

Keeping it simple: we're just going to look at the mean and the median of the seedings for the four teams in the Final Four. Median seems like it's probably more appropriate.

```{r get_data, echo=FALSE, message=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse)

df <- read_csv("data/final_four_seeds.csv") |> 
  pivot_longer(-year, names_to = "team", values_to = "team_seed") |> 
  mutate(seed = gsub("^.*\\#(.*)$", "\\1", team_seed)) |> 
  mutate(seed = as.numeric(seed))

df_grouped <- df |> 
  group_by(year) |> 
  summarize(seed_median = median(seed), 
            seed_mean = mean(seed),
            seed_max = max(seed),
            seed_min = min(seed)) |> 
  ungroup()

# Set up plotting the median and mean
df_mean_median <- df_grouped |> 
  select(year, seed_median, seed_mean) |> 
  pivot_longer(-year, names_to = "metric", values_to = "value")

```

Let's start by looking at just the raw data, though:

```{r echo=FALSE, message=FALSE}

ggplot(df, aes(x = year, y = seed)) +
  # geom_point(color = "#009CDE", alpha = 0.5) +
  geom_point(shape = 16, fill = "black", alpha = 0.2) +
  geom_line(aes(x = year, y = value, color = metric, size = metric), data = df_mean_median) +
  scale_y_reverse(limits = c(16,1),breaks = seq(16,1)) +
  scale_x_continuous(breaks = seq(1980, 2020, 5)) +
  scale_color_manual(values = c("gray70", "#009CDE"), labels = c("Mean Seed   ", "Median Seed  ")) +
   scale_size_manual(values = c(0.4,0.8), labels = c("Mean Seed   ", "Median Seed  ")) +
  labs(title = "NCAA Men's March Madness Final Four by Seed",
       subtitle = "Seeding of Each Individual Team Shown As a Dot",
       y = "Team Seed") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top",
        legend.title = element_blank())

```
